
The $\mathcal{S}$-operad framework naturally captures fundamental complex systems properties through its mathematical structure. We demonstrate how key features identified in complex systems literature emerge directly from operadic composition.

\subsection{Built-in support for structural properties}

The operadic structure provides native support for fundamental characteristics of complex systems.

\subsubsection{Structural: hierarchy, modularity and near-decomposability}

The $\mathcal{S}$-operad framework inherently captures the fundamental structural properties that define complex systems. \textbf{Near-decomposability} --- systems with stronger internal than external interactions --- emerges naturally from the operadic structure where each operation (box) defines a clear subsystem with well-defined input/output interfaces that enforce boundary conditions through composition rules. Strong internal interactions are captured by the temporal probability space within each operation, while weak external interactions are modeled by stochastic kernels between operations, creating the characteristic \textbf{interaction strength separation}. The framework supports \textbf{hierarchical organization} through nested composition where smaller operations combine into larger ones, with each level of composition corresponding to a distinct hierarchical level where higher-level operations operate on outputs of lower-level ones. Different hierarchical levels operate at different temporal scales through the filtration structure $\{\mathcal{F}_t\}$, allowing higher-level operations to have coarser temporal discretization than lower-level ones, naturally capturing \textbf{multi-scale dynamics}. This \textbf{time-scale separation} is enforced by the 2-step composition process that first forms independent products then applies dependencies, ensuring that subsystem dynamics remain distinguishable across scales.

\subsubsection{Modularity and compositionality}

Complex systems exhibit \textbf{modular organization} where components can be recombined in different configurations, a property that emerges directly from operadic composition principles. Each operation functions as a module with standardized interfaces (ports) that enable systematic construction of larger systems from smaller components through \textbf{operadic substitution}. The composition rules ensure that modules can only be connected when their interfaces match, preventing incompatible combinations while enabling flexible recombination of components. This \textbf{interface compatibility} creates a \textbf{compositional semantics} where the meaning (temporal dynamics) of a composite system is determined by the meanings of its components and their interaction patterns, following strict compositional principles that guarantee predictable system behavior from modular assembly.

\subsubsection{Scale-invariance and fractal structure}

\textbf{Scale-invariance} and \textbf{self-similar patterns} across scales emerge naturally from operadic self-composition, enabling the framework to capture the fractal-like organization characteristic of many complex systems. Operations can compose with themselves at different scales through \textbf{recursive application} $f \circ f \circ f \cdots$, creating hierarchical structures where the same interaction pattern repeats at multiple organizational levels. This self-similar composition generates the \textbf{recursive structures} observed in biological branching patterns, social network hierarchies, and economic market dynamics. The temporal probability spaces can exhibit \textbf{scale-free behavior} when stochastic kernels preserve \textbf{power-law distributions} during composition—if the initial distribution follows $\mathbb{P}(X) \sim X^{-\alpha}$, appropriately chosen kernels maintain this scaling relationship through successive compositions. This mathematical property enables the framework to model systems that lack \textbf{characteristic scales}, such as neural networks with scale-free connectivity or ecosystems with power-law species distributions. The operadic structure naturally supports \textbf{recursive definitions} where complex systems are defined in terms of smaller versions of themselves, generating the self-similar hierarchies that exhibit \textbf{statistical invariance} across observation scales.

\subsection{Emergence}

Unlike the structural properties examined above, emergence represents a fundamentally different challenge: identifying when and how higher-level properties arise that cannot be reduced to or predicted from individual component behaviors. The $\mathcal{S}$-operad framework provides a novel mathematical criterion for emergence based on the transition between explicit individual modeling and aggregate approximations.

\subsubsection{The emergence transition criterion}

Recall from our epidemic example the transition from explicit individual modeling (say N=3 where individual variations matter) to valid aggregate approximations (say N=100,000 where statistical behavior emerges). This transition point provides a precise mathematical criterion for emergence. We propose that emergence occurs when the composition of all individual stochastic kernels can be approximated by simpler aggregate kernels within controlled error bounds:

\[\left\| K_{macro} - \text{Proj} \left( \bigcirc_{i,j} K_{i \to j} \right) \right\| < \epsilon\]

where $K_{macro}$ operates on aggregate observables (like total infection rate $R(t) = \frac{1}{N} \sum_{i=1}^N \phi_i(t)$) and $\text{Proj}$ is the projection operator that maps individual-level dynamics to aggregate quantities.

Below the emergence threshold (small N) individual variations dominate and the approximation fails. The system behavior depends critically on specific individual properties, for example, a single super-spreader can dramatically alter global dynamics. No meaningful aggregate description exists. Above the emergence threshold (large N), individual variations become statistically insignificant compared to collective behavior. The aggregate observables follow predictable dynamics that are irreducible to individual properties where knowing everything about individuals still requires the aggregate calculation to predict system behavior.

\subsubsection{Dobrushin conditions and propagation of chaos}

Our emergence criterion connects directly to established results in statistical mechanics and probability theory. The transition from individual to aggregate dynamics mirrors the conditions for \textbf{Dobrushin uniqueness} \citep{dobrushin1968description}, which characterizes when correlations decay exponentially in lattice systems.

For a system with interdependence matrix $C$ where $C_{ij}$ measures how strongly the distribution at site $i$ depends on the configuration at site $j$, the \textbf{Dobrushin condition} states that emergence occurs when:

\begin{equation}
\|C\|_\infty = \max_i \sum_j C_{ij} < 1
\end{equation}

Under this condition, correlations decay exponentially with correlation length $\xi \sim 1/|\ln\|C\|_\infty|$. In our $\mathcal{S}$-operad framework, the kernels $K_{i \to j}$ naturally define such interdependence: $C_{ij}$ measures how strongly kernel $K_{i \to j}$ couples the stochastic processes at ports $i$ and $j$.

This connects to the \textbf{propagation of chaos} literature \citep{sznitman1991topics}, which provides quantitative bounds on when $N$-particle systems converge to mean-field limits. For systems with Lipschitz interaction kernels satisfying a confinement condition (interaction strength $\eta$ smaller than drift rate $c$), the Wasserstein distance between the $N$-particle empirical measure and the mean-field limit satisfies:

\begin{equation}
\mathbb{W}_1(\mu_t^N, \mu_t^{MF}) \leq C \cdot e^{-2(c-\eta)t} \cdot \mathbb{W}_1(\nu, \mu_0) + C \cdot \frac{\eta}{c-\eta} \cdot N^{-1/2}
\end{equation}

The fundamental $\mathcal{O}(N^{-1/2})$ convergence rate represents the \textbf{emergence threshold}: for large $N$, individual fluctuations become negligible compared to collective behavior. Modern results \citep{jabin2018quantitative} extend this to singular kernels, showing that emergence can occur even with strong local interactions if the system maintains sufficient global regularity.

In our operadic setting, this translates to the condition that pairwise correlations between components decay as $\text{Corr}(\phi_i, \phi_j) \sim N^{-\alpha}$ with $\alpha > 1/2$. When this holds, aggregate observables concentrate around their means faster than the standard CLT rate, making the macro-approximation $K_{macro}$ valid. Conversely, if $\alpha \leq 1/2$, correlations persist at all scales and emergence fails and the composed kernel $\bigcirc_{i,j} K_{i \to j}$ cannot be simplified.

\subsubsection{Network topology effects}

The emergence criterion depends critically on the topology of interactions between components. In our $\mathcal{S}$-operad framework, the network of stochastic kernels $K_{i \to j}$ determines how correlations decay:

\begin{itemize}
    \item \textbf{Sparse networks} with bounded degree facilitate emergence at moderate system sizes, as correlations decay quickly across limited connections
    \item \textbf{Dense networks} where each component influences many others require larger $N$ for emergence, as correlations persist through multiple pathways
    \item \textbf{Hierarchical networks} with community structure enable emergence at multiple scales, within communities first, then between communities
\end{itemize}

For our epidemic example, this explains why contact networks with community structure (strong local clustering, sparse inter-community connections) facilitate emergence at community scale, while well-mixed populations require larger sizes for meaningful aggregate descriptions.

This mathematical criterion reveals why certain macro descriptions work so well despite computational irreducibility claims \citep{bar2013computability}. While individual trajectories may be computationally irreducible and requiring full simulation, aggregate statistics become predictable above the emergence threshold. The macro-kernel $K_{macro}$ provides algorithmic shortcuts for predicting collective behavior even when individual behavior remains unpredictable. The resolution lies in \textbf{measure concentration} \citep{stanley1999scaling}. Above the emergence threshold, the vast majority of individual configurations yield similar aggregate behavior. The typical behavior becomes predictable through $K_{macro}$ even though specific individual trajectories remain computationally complex. Emergence occurs precisely when the information needed to predict aggregate dynamics becomes dramatically smaller than the information in individual states. The projection $\text{Proj}$ represents massive information compression that preserves predictive power for macro-observables while discarding individual details.

The criterion also reveals why certain properties can only be understood at the aggregate level and cannot be reduced to individual component properties. Consider the epidemic example above the emergence threshold. The infection rate dynamics $\dot{R}(t)$ depends on aggregate contact patterns, travel flows, and population mixing—properties that emerge from individual interactions but cannot be computed from individual properties alone. Even with complete knowledge of each person's immune state $\phi_i(t)$, predicting $\dot{R}(t)$ requires the full kernel composition $\bigcirc_{i,j} K_{i \to j}$, which is computationally equivalent to solving the aggregate problem. Above the emergence threshold, the system exhibits properties that have no individual-level analogues. Herd immunity thresholds, critical vaccination rates, and epidemic phase transitions \citep{pastor2015epidemic} are collective properties that emerge from the statistical structure of interactions but cannot be defined for individual agents. These represent genuinely new system properties.

Emergent properties acquire causal power in the sense that macro-states influence future macro-evolution through $K_{macro}$ in ways that cannot be reduced to sums of individual influences. The community infection state affects travel patterns, which affect other communities—a causal chain that operates at the emergent level.
